---
title: "Raster resampling in R"
author: "Cesar Aybar"
date: '2018-12-05'
output:
  html_document:
    toc: true
    toc_float: true
tags: ["GdalUtils", "raster", "furrr", "plotly"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## 1. Intro
A typical workflow for earth analysts always will have to do with several spatial data sources (e.g. Sentinel, MODIS and Landsat) each one with a particular extent and pixel resolution. **Resampling** is the technique used to *homogenize* these spatial attributes.

![Image obtained from [here](http://desktop.arcgis.com/en/arcmap/latest/extensions/spatial-analyst/performing-analysis/cell-size-and-resampling-in-analysis.htm)](http://desktop.arcgis.com/en/arcmap/latest/extensions/spatial-analyst/performing-analysis/GUID-F4B313D5-9585-4687-B5EE-37E9C9243EBE-web.png)

This post demonstrates how **easy** and **incredibly** fast is resampling gridded data in R using either the [raster](https://cran.r-project.org/web/packages/raster/index.html)
 and [gdalUtils](https://cran.r-project.org/web/packages/gdalUtils/index.html) packages. If you do not familiarize with they , first, look at this [tutorial](https://geoscripting-wur.github.io/IntroToRaster/) or run in R:
```{r vignette}
vignette("Raster", package = "raster") # Run me in R!
```

While it is true that the `raster` 'high-level' functions allow performing fast operation thanks to optimized back-end C code, some gridded manipulation techniques such as `raster::resample()` can become incredibly slow as the raster size increase. An effective alternative but a little more complicated to code is use a **wrapper library** (gdalUtils).

## 2. Wrapper libraries

In few words, a *wrapper* library (not confuse with a binding library) is a piece of code which translate a library's existing (e.g GDAL) into a different interface (e.g an R package). The main advantage is that you will use GDAL in its original language (C++).  Check it [this](https://stackoverflow.com/questions/8628326/what-is-the-difference-between-a-wrapper-bindings-and-a-port) for a more comprehensible explanation.

## 3. Example
Get the climatological gridded rainfall dataset [PISCOpclim](http://ons.snirh.gob.pe/Peru/maproom/Monitoring/Meteorological/PISCO-Prec-v20.pdf) (~10 km) for [Arequipa](https://en.wikipedia.org/wiki/Arequipa_Region) and
measure running time when changing progressively the pixel size.

### 3.1 Libraries
Before coding do not forget install and load the following packages.

#####Data Visualization libraries
```{r plot_libs, warning=FALSE, message=FALSE, eval = TRUE}
library('ggplot2')
library('plotly')
library('cptcity')
library('rasterVis')
library('mapview')
library('kableExtra')
```

#####R as a GIS
```{r GIS, warning=FALSE, message=FALSE, eval = TRUE}
library('gdalUtils')
library('sf')
library('raster')
```

#####Functional programming and tidyverse
```{r load_libs, warning=FALSE, message=FALSE, eval = TRUE}
library('purrr')
library('furrr')
library('tidyverse')
```

### 3.2 Downloading/Stacking PISCOpclim
```{r getdata, message = FALSE}
user <- 'publi_dgh2' # ftp user 
pass <- '123456'   # ftp user
ftp <- 'ftp.senamhi.gob.pe/PISCOp_V2.1_beta/' # url
dir <- 'PISCOp_climatology/PISCOp_climatology/' # urn
months <- sprintf('%02d', 1:12) # GEOtiff names

clim <- sprintf('ftp://%s:%s@%s%s%s.tif', user, pass, ftp, dir, months)

td <- tempdir() #temporal dir
tmp_files <- sprintf('%s%s', td, basename(clim))

ftp_n <- mapply(download.file, clim, tmp_files, quiet = TRUE)

clim_stack <- stack(tmp_files) %>% 
  'names<-'(month.abb) %>% '*'(1) # change the RasterLayer names
clim_stack[clim_stack <= 0] <- 0
storage.mode(clim_stack[]) = "integer" #round to an integer

mapply(file.remove, tmp_files)
```

### 3.2 Downloading/Uncompressing geometries

```{r down_geom, message = FALSE}
webpage_github <- 'https://github.com/csaybar/csaybar.github.io/'
file <- 'raw/master/data/post01.zip'
link <- sprintf('%s%s', webpage_github, file)
fn <- sprintf('%s.zip', tempfile())
download.file(link, fn)
td <- tempdir()
unzip(fn, ex = td)
```


### 3.3 Visualizing PISCOpclim

```{r viz, message = FALSE}

name_sud <- sprintf('%s/sud_peru_2.shp', td)
sud_sp <- st_read(name_sud, quiet = TRUE) %>% as('Spatial')

palette <- cpt('ukmo_wow_humidity') 
range <- seq(0,800,100)
levelplot(clim_stack,
          main = 'PISCOp Climatological Precipitation', 
          margin = T,    
          colorkey = list(
            labels = list(at = range, font = 4),
            axis.line = list(col = 'black')),    
          col.regions = palette,
          at = range) +
  layer(sp.polygons(sud_sp, col = rgb(1, 1, 1, 0.4), lwd = 1))
```

```{r img_01, echo = FALSE, eval = TRUE}
load('../../data/post01.Rdata')
sud_sp <- figures$fig_01$shp

palette <- cpt('ukmo_wow_humidity') 
range <- seq(0, 800, 100)
levelplot(figures$fig_01$stack,
          main = 'PISCOp Climatological Precipitation', 
          margin = TRUE,    
          colorkey = list(
            labels = list(at = range, font = 4),  
            axis.line = list(col = 'black')),    
          col.regions = palette,
          at = range) +
  layer(sp.polygons(sud_sp, col = rgb(1, 1, 1, 0.4), lwd = 1))
```



### 3.4 Arequipa Region
```{r, message = FALSE}
# Creating a polygon of the PISCOp extent
piscop_extent <- clim_stack[[1]] %>% 
                  extent %>% 
                  raster(nrows = 1, ncols = 1) %>% 
                  as('SpatialPolygons') %>% 
                  st_as_sf()

piscop_extent$NOMBRE = 'PISCOp extent'

name_arq <- sprintf('%s/arequipa_simply.shp', td)
arequipa <- st_read(name_arq, quiet = TRUE) %>% as('Spatial')

m_01 <- mapview(arequipa)
m_02 <- mapview(piscop_extent, fill = NA)

m_01 + m_02
```

```{r img_02, echo=FALSE, eval = TRUE}
figures$fig_02
```

### 3.5 Method 01: Resampling via `raster` package

```{r raster}
raster_base <- crop(clim_stack[[1]], arequipa)
res(raster_base) <- 0.05

clim_1_a <- raster::resample(clim_stack,
                             raster_base,
                             method='bilinear')

plot(clim_1_a[[1]], main = 'Jan Precipitation in Arequipa')
plot(arequipa, add = TRUE)
```

```{r img_03, echo = FALSE, eval = TRUE}
plot(figures$fig_03$rr, main = 'Jan Precipitation in Arequipa')
plot(figures$fig_03$shp, add = TRUE)
```


### 3.6 Method 02: Resampling via `gdalUtils` package

```{r gdalwarp}
#' Resampling a Raster* object via GDAL
#' 
#' @param r Raster* object to be resampled
#' @param r_base Raster* object with parameters that r
#' should be resampled to.
#' @param method Character. GDAL resampling_method
#' ("near"|"bilinear"|"cubic"|"cubicspline"|
#'  "lanczos"|"average"|"mode"|"max"|"min"|
#'  "med"|"q1"|"q3")
#'
#' @export
gdal_resample <- function(r, r_base, method = 'bilinear') {
  
  #Geometry attributes
  t1 <- c(xmin(r_base), ymin(r_base), 
          xmax(r_base), ymax(r_base))
  res <- res(r_base)
  
  #Temporal files
  tmp_outname <- sprintf('%s.tif', tempfile())
  tmp_inname <- sprintf('%s.tif', tempfile())
  writeRaster(r, tmp_inname)
  
  #GDAL time!
  gdalwarp(tmp_inname, tmp_outname, 
           tr = res, te = t1, r = method)
  resample_raster = raster(tmp_outname)
  
  return(resample_raster)
}
```

**Sequential approach**
```{r seq}
clim_1_b <- names(clim_stack) %>% 
  map(~ gdal_resample(clim_stack[[.x]], raster_base)) %>%
  stack()

plot(clim_1_b[[1]], main = 'Jan Precipitation in Arequipa')
plot(arequipa, add = TRUE)
```

```{r img_04, echo=FALSE, eval = TRUE}
plot(figures$fig_04$rr, main = 'Jan Precipitation in Arequipa')
plot(figures$fig_04$shp, add = TRUE)
```

**Multiprocessing approach**

```{r mypc, eval = TRUE}
message(sprintf('%s \n',system('lscpu',intern = T))) # only valid for linux users
```

```{r}
plan(multiprocess)
```

```{r}
clim_1_c <- names(clim_stack) %>% 
  future_map(~ gdal_resample(clim_stack[[.x]], raster_base)) %>%
  stack()

plot(clim_1_c[[1]], main = 'Jan Precipitation in Arequipa')
plot(arequipa, add = TRUE)
```

```{r img_05, echo=FALSE, eval = TRUE}
plot(figures$fig_05$rr, main = 'Jan Precipitation in Arequipa')
plot(figures$fig_05$shp, add = TRUE)
```

## 4. ``raster`` vs ``gdalUtils`` Comparison (Simple Benckmark)
```{r benckmark}
cell_size <- c(0.05, 0.025, 0.01, 0.005, 0.0025, 0.00175, 0.0015 , 0.001)

for(x in cell_size) { 
  
  raster_base <- crop(clim_stack[[1]], arequipa)
  res(raster_base) <- x

  time_m01 <- system.time(
    clim_1_a <- resample(clim_stack,
                         raster_base,
                         method='bilinear'))

  time_m02 <- system.time(
    clim_1_b <- names(clim_stack) %>% 
      map(~ gdal_resample(clim_stack[[.x]], raster_base)) %>%
      stack())

  time_m03 <- system.time(
    clim_1_c <- names(clim_stack) %>% 
      future_map(~ gdal_resample(clim_stack[[.x]],
                                 raster_base)) %>%
      stack())
  
  df_size <- data_frame(
    pixel_size = x,
    npixels = ncell(clim_1_a)*12,
    raster = time_m01[3],
    gdalUtils = time_m02[3],
    gdalUtils_4par = time_m03[3]
  )
  if (x == 0.05) {
    df_base <- df_size
  } else {
    df_base <- bind_rows(df_base, df_size)
  }
}
```

```{r,eval = TRUE , echo = FALSE}
load('../../data/post01.Rdata')
df_base <- figures$fig_06
kable(df_base)%>%
  kable_styling(bootstrap_options = "striped",
                full_width = F)
```

## 5. Results
The bellow plot shows a considerable improvement when using **gdaltUtils** respect to **raster** package only for cell sizes < 0.005 km. (50M > pixels). **gdalUtils** using a multiprocessing approach (and 4 cores) was able to generate more than **100M of pixels** in less than 12 seg!.

```{r,eval=T,message = FALSE}
df_base %>%  plot_ly(x=~npixels) %>% 
  add_trace(y = ~raster, name = 'raster',
            mode = 'lines+markers') %>%
  add_trace(y = ~gdalUtils, name = 'GdalUtils',
            mode = 'lines+markers') %>%
  add_trace(y = ~gdalUtils_4par, name = 'GdalUtils_par',
            mode = 'lines+markers') %>% 
  layout(xaxis = list(type = "log",title = 'nPixels'),
         yaxis = list(title = 'time (s)'))
```

## 6. Epilogue
I hope you have enjoyed reading this post. If you are interested in still continuing increasing the speed of your spatial processings maybe you want to run the following in your R console:

```{r vignette_02}
vignette("functions", package = "raster") # Run me in R!
rasterOptions()
```

Oh, and before I forget. At the near future, the [terra](https://github.com/rspatial/terra?fbclid=IwAR0gEKvR3AxrvtpyxblIe-jXC2RsZg85IhlI2wDHvGYncOozVdWSMncsoEI) package will replacement the raster package. The first (alpha) release is expected by January 2019. I promise will make an update of this post after the package's release.
